### 2020-09-28 priority-gv.tex
Added small bits of text.

Session types are annotated with \emph{priorities} $\cs{o}\in\mathbb{N}$.
The types $\ty{\tysend[\cs{o}]{T}{S}}$ and $\ty{\tyrecv[\cs{o}]{T}{S}}$ type the endpoints of a channel over which we send or receive a value of type $\ty{T}$, and then proceed as $\ty{S}$.
The types $\ty{\tyends[\cs{o}]}$ and $\ty{\tyendr[\cs{o}]}$ type endpoints of a channel whose session has finished, and over which we send or receive a ping before closing the channel.
\[
\begin{array}{lcl}
  \ty{S}
  & \coloneqq & \ty{\tysend[\cs{o}]{T}{S}}
    \sep        \ty{\tyrecv[\cs{o}]{T}{S}}
    \sep        \ty{\tyends[\cs{o}]}
    \sep        \ty{\tyendr[\cs{o}]}
\end{array}
\]
Value types are based on the types of the linear $\lambda$-calculus.
The types $\ty{\typrod{T}{U}}$, $\ty{\tyunit}$, $\ty{\tysum{T}{U}}$, and $\ty{\tyvoid}$ are the standard linear $\lambda$-calculus product type, unit type, sum type, and empty type.
The type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ is the standard linear function type, annotated with \emph{priority bounds} $\cs{p},\cs{q}\in\mathbb{N}\cup\{\cs{\pbot},\cs{\ptop}\}$.
Session types are value types, as endpoints are values.
\[
\begin{array}{lcl}
  \ty{T}, \ty{U}
  & \coloneqq & \ty{\typrod{T}{U}}
    \sep        \ty{\tyunit}
    \sep        \ty{\tysum{T}{U}}
    \sep        \ty{\tyvoid}
    \sep        \ty{\tylolli[\cs{p},\cs{q}]{T}{U}}
    \sep        \ty{S}
\end{array}
\]
Given a function with type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$, the priority bound $\cs{p}$ is a \emph{lower bound} on the priorities of the channel names captured by the body of the function, and the priority bound $\cs{q}$ is an \emph{upper bound} on the priority of the communications that take place as a result of applying the function.
We discuss this in more detail when we give the typing rules.
We define syntactic sugar for the type of \emph{pure functions}.
\[
  \ty{\tylolli{T}{U}} \elabarrow \ty{\tylolli[\cs{\ptop},\cs{\pbot}]{T}{U}}
\]
A typing context $\ty\Gamma$ is a partial function associating names to types.
\[
\begin{array}{lcl}
  \ty{\Gamma}, \ty{\Delta}
  & \coloneqq & \ty{\emptyenv}
    \sep        \ty{\Gamma}, \tmty{x}{T}
\end{array}
\]
\[
\begin{array}{lcl}
  \ty{\co{\tysend[\cs{o}]{T}{S}}} & = & \ty{\tyrecv[\cs{o}]{T}{\co{S}}} \\
  \ty{\co{\tyrecv[\cs{o}]{T}{S}}} & = & \ty{\tysend[\cs{o}]{T}{\co{S}}}
\end{array}
\qquad
\begin{array}{lcl}
  \ty{\co{\tyends[\cs{o}]}} & = & \ty{\co{\tyendr[\cs{o}]}} \\
  \ty{\co{\tyendr[\cs{o}]}} & = & \ty{\co{\tyends[\cs{o}]}}
\end{array}
\]
\[
\begin{array}{lcl}
  \pr(\ty{\tysend[\cs{o}]{T}{S}})            & = & \cs{o} \\
  \pr(\ty{\tyrecv[\cs{o}]{T}{S}})            & = & \cs{o} \\
\end{array}
\qquad
\begin{array}{lcl}
  \pr(\ty{\tyends[\cs{o}]})                  & = & \cs{o} \\
  \pr(\ty{\tyendr[\cs{o}]})                  & = & \cs{o} \\
\end{array}
\]
\[
\begin{array}{lcl}
  \minpr(\ty{\typrod{T}{U}})                 & = & \minpr({\ty{T}})\sqcap\minpr({\ty{U}}) \\
  \minpr(\ty{\tysum{T}{U}})                  & = & \minpr({\ty{T}})\sqcap\minpr({\ty{U}}) \\
  \minpr(\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}) & = & \cs{p} \\
  \\
  \minpr(\ty{\Gamma}, \tmty{x}{A})           & = & \minpr(\ty{\Gamma})\sqcap\pr(\ty{A})
\end{array}
\qquad
\begin{array}{lcl}
  \minpr(\ty{\tyunit})                       & = & \ptop \\
  \minpr(\ty{\tyvoid})                       & = & \ptop \\
  \minpr(\ty{S})                             & = & \pr(\ty{S}) \\
  \\
  \minpr(\ty{\emptyenv})                     & = & \ptop
\end{array}
\]

\paragraph*{Static Term Syntax}
We use $\tm{x}$, $\tm{y}$, $\tm{z}$, and $\tm{w}$ to range over variable names. Occasionally, we use $\tm{a}$, $\tm{b}$, and $\tm{c}$ to range over \emph{free} channel endpoints.
\[
\begin{array}{lcl}
  \tm{L}, \tm{M}, \tm{N}
  & \coloneqq & \tm{x}
    \sep        \tm{K}
    \sep        \tm{\lambda x.M}
    \sep        \tm{M\;N} \\
  & \sep      & \tm{\unit}
    \sep        \tm{\letunit{M}{N}} \\
  & \sep      & \tm{\pair{M}{N}}
    \sep        \tm{\letpair{x}{y}{M}{N}} \\
  & \sep      & \tm{\inl{M}}
    \sep        \tm{\inr{M}}
    \sep        \tm{\casesum{L}{x}{M}{y}{N}} \\
  & \sep      & \tm{\absurd{M}} \\
  \tm{K}
  & \coloneqq & \tm{\link}
    \sep        \tm{\new}
    \sep        \tm{\spawn}
    \sep        \tm{\send}
    \sep        \tm{\recv}
    \sep        \tm{\wait}
    \sep        \tm{\close}
\end{array}
\]

>>>>

Session types are annotated with \emph{priorities} $\cs{o}\in\mathbb{N}$.
The types $\ty{\tysend[\cs{o}]{T}{S}}$ and $\ty{\tyrecv[\cs{o}]{T}{S}}$ type the endpoints of a channel over which we send or receive a value of type $\ty{T}$, and then proceed as $\ty{S}$.
The types $\ty{\tyends[\cs{o}]}$ and $\ty{\tyendr[\cs{o}]}$ type endpoints of a channel whose session has finished, and over which we send or receive a ping before closing the channel.
\[
\begin{array}{lcl}
  \ty{S}
  & \coloneqq & \ty{\tysend[\cs{o}]{T}{S}}
    \sep        \ty{\tyrecv[\cs{o}]{T}{S}}
    \sep        \ty{\tyends[\cs{o}]}
    \sep        \ty{\tyendr[\cs{o}]}
\end{array}
\]
Value types are based on the types of the linear $\lambda$-calculus.
The types $\ty{\typrod{T}{U}}$, $\ty{\tyunit}$, $\ty{\tysum{T}{U}}$, and $\ty{\tyvoid}$ are the standard linear $\lambda$-calculus product type, unit type, sum type, and empty type.
The type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ is the standard linear function type, annotated with \emph{priority bounds} $\cs{p},\cs{q}\in\mathbb{N}\cup\{\cs{\pbot},\cs{\ptop}\}$.
Session types are value types, as endpoints are values.
\[
\begin{array}{lcl}
  \ty{T}, \ty{U}
  & \coloneqq & \ty{\typrod{T}{U}}
    \sep        \ty{\tyunit}
    \sep        \ty{\tysum{T}{U}}
    \sep        \ty{\tyvoid}
    \sep        \ty{\tylolli[\cs{p},\cs{q}]{T}{U}}
    \sep        \ty{S}
\end{array}
\]
Given a function with type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$, the priority bound $\cs{p}$ is a \emph{lower bound} on the priorities of the channel names captured by the body of the function, and the priority bound $\cs{q}$ is an \emph{upper bound} on the priority of the communications that take place as a result of applying the function.
We discuss this in more detail when we give the typing rules.
We define syntactic sugar for the type of \emph{pure functions}.
\[
  \ty{\tylolli{T}{U}} \elabarrow \ty{\tylolli[\cs{\ptop},\cs{\pbot}]{T}{U}}
\]
A typing context $\ty\Gamma$ is a partial function associating names to types.
\[
\begin{array}{lcl}
  \ty{\Gamma}, \ty{\Delta}
  & \coloneqq & \ty{\emptyenv}
    \sep        \ty{\Gamma}, \tmty{x}{T}
\end{array}
\]
Duality plays a crucial role in session types. The two endpoints of a channel are assigned dual types, which ensures that, for instances, whenever one program \emph{sends} a value on a channel, the program on the other end is waiting to \emph{receive}. Each \emph{session type} $\ty{T}$ has a dual, written $\ty{\co{T}}$. Duality is an involutive function which preserves priorities.
\[
\begin{array}{lcl}
  \ty{\co{\tysend[\cs{o}]{T}{S}}} & = & \ty{\tyrecv[\cs{o}]{T}{\co{S}}} \\
  \ty{\co{\tyrecv[\cs{o}]{T}{S}}} & = & \ty{\tysend[\cs{o}]{T}{\co{S}}}
\end{array}
\qquad
\begin{array}{lcl}
  \ty{\co{\tyends[\cs{o}]}} & = & \ty{\co{\tyendr[\cs{o}]}} \\
  \ty{\co{\tyendr[\cs{o}]}} & = & \ty{\co{\tyends[\cs{o}]}}
\end{array}
\]
The function $\pr$ returns the smallest priority of a session type. The type system guarantees that the top-most connective always holds the smallest priority, so we simply return the priority of the top-most connective.
\[
\begin{array}{lcl}
  \pr(\ty{\tysend[\cs{o}]{T}{S}})            & = & \cs{o} \\
  \pr(\ty{\tyrecv[\cs{o}]{T}{S}})            & = & \cs{o} \\
\end{array}
\qquad
\begin{array}{lcl}
  \pr(\ty{\tyends[\cs{o}]})                  & = & \cs{o} \\
  \pr(\ty{\tyendr[\cs{o}]})                  & = & \cs{o} \\
\end{array}
\]
We extend the function $\pr$ to value types and typing contexts by returning the smallest priority in the type or context, or $\cs{\top}$ if there is no priority.
\todo{explain function type.}
\[
\begin{array}{lcl}
  \minpr(\ty{\typrod{T}{U}})                 & = & \minpr({\ty{T}})\sqcap\minpr({\ty{U}}) \\
  \minpr(\ty{\tysum{T}{U}})                  & = & \minpr({\ty{T}})\sqcap\minpr({\ty{U}}) \\
  \minpr(\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}) & = & \cs{p} \\
  \\
  \minpr(\ty{\Gamma}, \tmty{x}{A})           & = & \minpr(\ty{\Gamma})\sqcap\pr(\ty{A})
\end{array}
\qquad
\begin{array}{lcl}
  \minpr(\ty{\tyunit})                       & = & \ptop \\
  \minpr(\ty{\tyvoid})                       & = & \ptop \\
  \minpr(\ty{S})                             & = & \pr(\ty{S}) \\
  \\
  \minpr(\ty{\emptyenv})                     & = & \ptop
\end{array}
\]

\paragraph*{Static Term Syntax}
Let $\tm{x}$, $\tm{y}$, $\tm{z}$, and $\tm{w}$ to range over variable names. Occasionally, we use $\tm{a}$, $\tm{b}$, $\tm{c}$, and $\tm{d}$ to range over \emph{free} channel endpoints. The term language is the linear $\lambda$-calculus, extended with constants for the communication primitives.
\[
\begin{array}{lcl}
  \tm{L}, \tm{M}, \tm{N}
  & \coloneqq & \tm{x}
    \sep        \tm{K}
    \sep        \tm{\lambda x.M}
    \sep        \tm{M\;N} \\
  & \sep      & \tm{\unit}
    \sep        \tm{\letunit{M}{N}} \\
  & \sep      & \tm{\pair{M}{N}}
    \sep        \tm{\letpair{x}{y}{M}{N}} \\
  & \sep      & \tm{\inl{M}}
    \sep        \tm{\inr{M}}
    \sep        \tm{\casesum{L}{x}{M}{y}{N}} \\
  & \sep      & \tm{\absurd{M}} \\
  \tm{K}
  & \coloneqq & \tm{\link}
    \sep        \tm{\new}
    \sep        \tm{\spawn}
    \sep        \tm{\send}
    \sep        \tm{\recv}
    \sep        \tm{\wait}
    \sep        \tm{\close}
\end{array}
\]
